Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments

### 2016-05-10 revised_submission/MSH.tex
Updated to revised version

%% Group authors per affiliation:
\author{R. Myhill, D. J. Frost}
\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany}
\cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
\ead{myhill.bob@gmail.com}

@@

\begin{abstract}
Hydrous melting at high pressures affects the physical properties, dynamics and chemical differentiation of the Earth. However, probing the compositions of hydrous melts at the conditions of the mantle transition zone has traditionally been challenging. In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were created using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the T-$X_{H_2O}$ liquidus curve develops an increasingly pronounced concave-down topology. The melting point reduction of enstatite and stishovite at low water contents exceeds that predicted by simple ideal models of hydrogen speciation. We discuss the implications of this work on the behaviour of melts in the deep upper mantle and transition zone, and present new models describing the partitioning of water between the olivine polymorphs and associated hydrous melts. 
\end{abstract}

@@

\section{Introduction}
Hydrous melts have a remarkable influence on Earth's surface and interior. Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes. The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}. This deep water cycle must extend to at least 500 km depth, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}. In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage. It has also been argued that neutral or negatively buoyant water-rich melts are to blame for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.

Processes involving deep-seated hydrous melts are controlled to a great extent by melt composition. Most importantly, the equilibrium proportion of H$_2$O in melts controls the extent of partial melting for a given bulk H$_2$O content. Together with solid-solid-melt dihedral angles, the extent of partial melting governs the permeability and therefore mobility of melt, and the ability to detect it via body wave seismology and electrical conductivity measurements.  Composition is also a key variable in determining melt density and viscosity. Finally, the change in equilibrium melt compositions with pressure and temperature has a significant influence on melt channelisation, which reduces melt-solid interaction and increases ascent rates.

Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure. Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth. Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O. As pressure increases, immiscibility between fluid and melt decreases, and the thermal stability of hydrous phases increases. As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one. Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.

Constraining melt compositions remains a major challenge in high-pressure experimental petrology. The primary problem is that water-rich melts are unquenchable. During quench, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour on decompression and sample preparation. The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}. Mass balance alone depends on knowing the volume proportions of solid phases and quench phases, being able to recognise and subtract quench overgrowths where they exist, and knowing the effective ambient-pressure density of the melt. All of these are associated with significant uncertainties. Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit would imply a large water content). However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam. For this reason, defocused beam techniques should probably be restricted to estimating non-volatile element ratios (such as Mg:Si).

Bracketing the liquidus requires only that the bulk composition during the experiment is known. This technique therefore seems a good alternative to estimating melt compositions ex-situ. The only problem is that at slab or ambient mantle temperatures, water contents in melts at high pressure are very high. Making up solid starting compositions with more water than a brucite-quartz mix has traditionally been avoided. Existing studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form if the sample chambers are sufficiently large \citep[e.g.][]{HM2012}. Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible. In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$), as a novel water source. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. It breaks down at higher temperatures, releasing four molecules of H$_2$O and forming platinum (IV) oxide PtO$_2$. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments.

\section{Experimental and analytical methods}
Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (H$_2$Pt(OH)$_6$). Before weighing, quartz was dried at 1000 $^{\circ}$C for 12 hours, while brucite was heated to 250 $^{\circ}$C, also over 12 hours. Both powders were stored in a desiccating oven at 130 $^{\circ}$C. Platinic acid is hygroscopic, so was stored in a vacuum desiccator. Powders were weighed and ground dry in an agate mortar for 30 minutes, using a mask, goggles and gloves to avoid physical contact with the platinic acid. Starting compositions were stored in glass vials in a vacuum desiccator. Compositions are listed in Table \ref{table:compositions}.

@@

Multianvil experiments were performed in the 5000 tonne press at the Bayerisches Geoinstitut (BGI). Cr-doped MgO octahedral multianvil assemblies with 18 mm edge length were used (Figure \ref{fig:assembly}). Two capsules were loaded into each assembly, with the open ends of the chambers facing each other and separated by six 0.05 mm thick Pt$_{90}$Rh$_{10}$ or Au foils. The assemblies were compressed to 13 GPa over four hours between eight tungsten carbide cubes with trucations of edge-length 11 mm. Pressure calibrations and details of the press can be found in \cite{FPTLDR2004} and \cite{KF2005}. The assemblies were then resistively heated via a stepped LaCrO$_3$ heater. The temperature was recorded using a W$_{97}$Re$_{3}$--W$_{75}$Re$_{25}$ (Type D) thermocouple inserted axially into the assembly. To avoid water loss, higher temperature runs were heated for a shorter duration. The experiments were quenched by cutting power to the heater. The assemblies were then decompressed to room pressure over 1000 minutes.

@@

\section{Modelling water solubility in melts}
\subsection{Background}
Oxygen in neutral MSH melt species can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group. The general equation describing the reaction between melt species is
\begin{equation}
\textrm{H}_2\textrm{O} + \textrm{O}_{br}^{2-} \rightleftharpoons 2 \textrm{OH}_{tm}^-
\label{eqn:speciation}
\end{equation}

@@

These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers. A high proportion of monomers and dimers exist in relatively dilute solutions, but not in the concentrated solutions investigated in this study.

A number of models have been presented to investigate the behaviour of concentrated hydrous melts. \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation. For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K_{(1)}=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers. No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment. Fixing $r$ places an implicit constraint on the maximum value of $K_{(1)}$ in water-rich compositions. For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K_{(1)}<\infty$. In practise, the equilibrium constant $K_{(1)}$ has been shown to be dependent on temperature and composition. \cite{SK1995} obtained a good fit to isochemical data with an expression of the form
\begin{equation}
\Delta G_{(1)} = a + b\,T
\end{equation}

@@

\subsection{The anhydrous melt model}
Thermodynamic models of MgO-SiO$_2$ melts are currently in their infancy. Many of the techniques used to constrain the thermodynamic properties (particularly the equations of state) of solids cannot be used on liquids as a result of their highly dynamic behaviour. Therefore in this study, we use MgO and SiO$_2$ melt models derived from molecular dynamics simulations \citep{DKS2013}, and fit a simplified mixing curve based on experimental data. First, however, the thermodynamic properties of the endmember melts need to be adjusted to fit the experimentally derived thermodynamic data at the melting point. This step was unnecessary in the original study of \cite{DKS2013} because solid properties were also obtained by molecular dynamics simulations. Such simulations do not provide good estimates of absolute energies and entropies, and so the values of the solids were obtained by fixing the melting point at a given pressure to the experimentally derived values (0 GPa, 3070 K for periclase, 14 GPa, 3120 K for stishovite). It turns out that the MD-determined entropy of periclase is almost identical to the experimentally determined values at all pressures and temperatures \citep{SLB2011}, so we only needed to adjust the standard state free energy. This leads to a periclase melting point of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}. We notes that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}. Current thermodynamic models for periclase cannot be extrapolated to the extreme temperatures predicted by the \cite{ZF2008} data. However, assuming that the volume change of melting is roughly constant with temperature (1.8 cm$^3$/mol at 13 GPa), the entropy of melting at a melting point of 5373 K would be about 50\% of that for a melting point of 4180 K (13.25 vs. 27.5 J/K/mol). At 13 GPa and 2000 K, MgO melt would actually be 15 kJ/mol \emph{more} stable as a result of this lower entropy.

The MD-determined entropy of stishovite is significantly lower than the experimentally determined value, so we increased the standard state entropy of the melt by 16.5 J/K/mol. Conveniently, we also have a way to test the volumes predicted by the model of \cite{DKS2013} at around 13 GPa. The experimentally-derived slope of the coesite melting curve is essentially zero at the triple point with stishovite at 13.7 GPa \citep{ZLGHF1993}. Therefore, from the Clausius-Clapeyron relationship $dT/dP = \Delta V / \Delta S$ the volumes of coesite and SiO$_2$ melt must be zero at this point. Unfortunately, the MD-derived SiO$_2$ melt model has its density crossover with coesite about 4 GPa lower. Therefore, we adjusted the bulk modulus and reference pressure so satisfy this constraint, and the volume at zero pressure. The adjustments required here may not be a failing of the MD simulations; it is likely that the coordination changes in the SiO$_2$ melt require a higher order Birch Murnaghan EoS than that used by \citep{DKS2013}. Our adjusted model is probably only accurate in a range of about $\pm$5 GPa around the triple point. The experimental data can also be used to judge the accuracy of the entropy of the SiO$_2$ melt model. The MD-derived entropy of melting is $\sim$40 J/K/mol at the triple point. This value is in perfect agreement with the value derived from the Clapeyron slope of the stishovite liquidus \citep{SL1995}. We note that the slope recently proposed by \cite{Millotetal2015} is much shallower, but their slope is a two parameter Simon's equation fit based on the triple point temperature at 13.7 GPa and the melting point at 500 GPa, and fits the data at 15--100 GPa \citep{SL1995, LAM1983} rather poorly. A Simon's equation fit is probably inappropriate over such extreme pressure ranges, especially where the melting point may be rather sensitive to changes in cation coordination.

A subregular solution model was adopted to describe the free energy of intermediate-composition MgO-SiO$_2$ melts. The two interaction parameters W$_{MS}$ and W$_{SM}$ at 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions \citep{PG1990, PW1993}. We assumed an ideal entropic contribution to mixing. The pressure dependencies were fit to the melting curves. It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower. However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).

@@

\begin{eqnarray}
p' = \frac{p_{H_2O}}{p_{H_2O}  + r \left(1.-p_{H_2O} \right)} \\
X_O = 1 - p' - \frac{ 0.5 - \sqrt{ 0.25 - \frac{K_{(1)}-4}{K_{(1)}} \left(p'-p'^2 \right)}}{\frac{K_{(1)}-4}{K_{(1)}} }\\
X_{H_2O} = X_O + 2 p' - 1 \\
X_{OH} = 1 - X_O - X_{H_2O}
\end{eqnarray}
\noindent $p'$ is the molar proportion of H$_2$O where the silicate is normalised on a one oxygen basis (rather than a one-cation basis). We assume that all bridging oxygens are equally likely to be protonated. The number of oxygens per formula unit $r$ thus increases monotonously from 1 to 2 across the MgO-SiO$_2$ binary ($r_{\textrm{per}}$ = 1, $r_{\textrm{fo, wad, rw}}$ = 4/3, $r_{\textrm{oen, cen}}$ = 3/2, and $r_{\textrm{coe, stv}}$ = 2). The ideal activities of the anhydrous and water components can then be calculated:
\begin{eqnarray}
a^{\textit{s}}_{\textit{ideal melt}} =X_O^r \\
a^{H_2O}_{\textit{ideal melt}} = X_{H_2O}
\end{eqnarray}

@@

\clearpage
\section{Results}
\subsection{Ex-situ observations}
Experimental details and results are shown in Table \ref{table:experiments}. Superliquidus runs revealed networks of dendritic crystals of brucite, forsterite, enstatite and stishovite. Below the liquidus, solid phases tended to aggregate at the cold end of the capsules. Periclase formed equant crystals often associated with quench overgrowths which appear brighter in BSE images. Forsterite formed tabular crystals about 100 $\mu$m long. Enstatite crystals were much smaller, forming ca. 5 $\mu$m equant grains. Stishovite also formed equant to blocky grains, but crystals only separated from the melts in chambers with a high water content. In addition to the difference in texture and shape between quench crystals liquidus phases, subliquidus phases typically contained many small inclusions of Pt-Rh oxides. A typical set of run products is shown in Figure \ref{fig:sem}.

\begin{figure}[h!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/assembly}
  \caption{The 18/11 octahedral assembly design used in this study.}
  \label{fig:assembly}
\end{figure}

>>>>

%% Group authors per affiliation:
\author{R. Myhill}
\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany \\ now at School of Earth Sciences, University of Bristol, Wills Memorial Building, Queens Road, Bristol BS8 1RJ, United Kingdom}
\cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
\ead{myhill.bob@gmail.com}

\author{D. J. Frost}
\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany}

@@

\begin{abstract}
Hydrous melting at high pressures affects the physical properties, dynamics and chemical differentiation of the Earth. However, probing the compositions of hydrous melts at the conditions of the deeper mantle such as the transition zone has traditionally been challenging. In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were created using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the T-$X_{H_2O}$ liquidus curve develops an increasingly pronounced concave-up topology. The melting point reduction of enstatite and stishovite at low water contents exceeds that predicted by simple ideal models of hydrogen speciation. We discuss the implications of these results with respect to the behaviour of melts in the deep upper mantle and transition zone, and present new models describing the partitioning of water between the olivine polymorphs and associated hydrous melts. 
\end{abstract}

@@

\section{Introduction}
Hydrous melts have a remarkable influence on Earth's surface and interior. Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes. The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}. This deep water cycle must extend to at least 500 km depth, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}. In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage. It has also been argued that neutral or negatively buoyant water-rich melts are responsible for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.

Processes involving deep-seated hydrous melts are controlled to a great extent by melt composition. Most importantly, the equilibrium proportion of H$_2$O in melts controls the melt fraction for a given bulk H$_2$O content. Together with solid-solid-melt dihedral angles, the extent of partial melting governs the permeability and therefore mobility of melt, and the ability to detect it via body wave seismology and electrical conductivity measurements. Melt densities and viscosities are also governed by their composition. Finally, changes in equilibrium melt composition with pressure, temperature and bulk composition can enhance melt channelisation, which reduces melt-solid interaction and increases ascent rates \citep[e.g.][]{SKA2001,WK2012}.

Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure. Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth \citep{Hirschmann2006}. Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O. As pressure increases, immiscibility between fluid and melt decreases \citep{XXXX}, and the thermal stability of hydrous phases increases \citep{XXXX}. As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one. Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.

Constraining melt compositions remains a major challenge in high-pressure experimental petrology. The primary problem is that water-rich melts are unquenchable. During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour on decompression and during sample preparation. The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}. Mass balance alone depends on being able to accurately estimate the volume proportions of solid phases and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt. All of these are associated with significant uncertainties. Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit implies a large water content). However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam. For this reason, defocused beam techniques should probably be restricted to estimating non-volatile element ratios (such as Mg:Si).

Here we use the approach of bracketing the liquidus surface to determine melt composition as a function of temperature in the system MgO-SiO$_2$-H$_2$O (MSH). This technique requires knowledge of the starting composition, techniques to effectively minimise water loss from the capsule and an ability to recognise the difference between liquidus and quench phases. We use this technique to develop thermodynamic models for water solubility in melts, and apply these to models of water partitioning between mantle phases and melts. We then apply these models to better understand hydrous melting under transition zone conditions.
%XXXX what, in general, have you done in this paper?

\section{Experimental and analytical methods}
Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (hexahydroxyplatinate(IV); H$_2$Pt(OH)$_6$). We use platinic acid to increase water contents to those expected of hydrous melts in equilibrium with mantle phases under subduction/convecting mantle conditions. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments. Before weighing, quartz was dried at 1000 $^{\circ}$C for 12 hours, while brucite was heated to 250 $^{\circ}$C, also over 12 hours. Both powders were stored in a desiccating oven at 130 $^{\circ}$C. Platinic acid is hygroscopic, so was stored in a vacuum desiccator. Powders were weighed and ground dry in an agate mortar for 30 minutes, using a mask, goggles and gloves to avoid physical contact with the platinic acid. Prepared powders were stored in glass vials in a vacuum desiccator until use.

We chose starting compositions to cover the brucite-quartz-water subregion of the MSH ternary. Chosen compositions include points on the binaries between water and brucite, forsterite, enstatite and stishovite, and a few additional compositions to constrain the locations of cotectics (Table \ref{table:compositions}).

@@

Multianvil experiments were performed in the 5000 tonne press at the Bayerisches Geoinstitut (BGI). Cr-doped MgO octahedral multianvil assemblies with 18 mm edge length were used (Figure \ref{fig:assembly}). Two capsules were loaded into each assembly, with the open ends of the chambers facing each other and separated by six 0.05 mm thick Pt$_{90}$Rh$_{10}$ or Au foils. The assemblies were compressed to 13 GPa over four hours between eight tungsten carbide cubes with trucations of edge-length 11 mm. Pressure calibrations and details of the press can be found in \cite{FPTLDR2004} and \cite{KF2005}. The assemblies were then resistively heated via a stepped LaCrO$_3$ heater. The temperature was recorded using a W$_{97}$Re$_{3}$--W$_{75}$Re$_{25}$ (Type D) thermocouple inserted axially into the assembly. To avoid water loss, higher temperature runs were heated for a shorter duration \citep[see][]{NDMPMF2016}. The experiments were quenched by cutting power to the heater. The assemblies were then decompressed to room pressure over 1000 minutes.

@@

\section{Modelling water solubility in melts}
\subsection{Background}
Oxygen in neutral MSH melt species exists either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group (terminal oxygens would require Si=O double bonds, which are extremely unstable). The general equation describing the reaction between melt species is therefore
\begin{equation}
\textrm{H}_2\textrm{O} + \textrm{O}_{br}^{2-} \rightleftharpoons 2 \textrm{OH}_{tm}^-
\label{eqn:speciation}
\end{equation}

@@

These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers. A high proportion of monomers and dimers exist in relatively dilute solutions \citep{XXXX}, but not in the concentrated solutions investigated in this study.

A number of models have been presented to investigate the behaviour of water-bearing silicate melts. \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation. For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K_{(1)}=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers. No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment. Fixing $r$ places an implicit constraint on the maximum value of $K_{(1)}$ in water-rich compositions. For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K_{(1)}<\infty$. In practise, the equilibrium constant $K_{(1)}$ has been shown to be dependent on temperature and composition. \cite{SK1995} obtained a good fit to isochemical data with an expression of the form
\begin{equation}
\Delta G_{(1)} = a + b\,T
\end{equation}

@@

\subsection{The anhydrous melt model}
\subsubsection{Endmembers}
Thermodynamic models of MgO-SiO$_2$ melts are currently in their infancy. Many of the techniques used to constrain the thermodynamic properties (particularly the equations of state) of solids cannot be used on liquids as a result of their highly dynamic behaviour. Therefore in this study, we use MgO and SiO$_2$ melt models derived from molecular dynamics simulations \citep{DKS2013}, and fit a simplified mixing curve based on experimental data. First, however, the thermodynamic properties of the endmember melts need to be adjusted to fit the experimentally derived thermodynamic data at the melting point. This step was unnecessary in the original study of \cite{DKS2013} because solid properties were also obtained by molecular dynamics simulations. Such simulations do not provide good estimates of absolute energies and entropies, and so the values of the solids were obtained by fixing the melting point at a given pressure to the experimentally derived values (0 GPa, 3070 K for periclase, 14 GPa, 3120 K for stishovite). It turns out that the MD-determined entropy of periclase is almost identical to the experimentally determined values at all pressures and temperatures \citep{SLB2011}, so we only needed to adjust the standard state free energy. This leads to a periclase melting point of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}. We note that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}. Current thermodynamic models for periclase cannot be extrapolated to the extreme temperatures predicted by the \cite{ZF2008} data. However, assuming that the volume change of melting is roughly constant with temperature (1.8 cm$^3$/mol at 13 GPa), the entropy of melting at a melting point of 5373 K would be about 50\% of that for a melting point of 4180 K (13.25 vs. 27.5 J/K/mol). At 13 GPa and 2000 K, MgO melt would actually be 15 kJ/mol \emph{more} stable as a result of this lower entropy.

The MD-determined entropy of stishovite is significantly lower than the experimentally determined value \citep{SLB2011}, so we increased the standard state entropy of the melt by 16.5 J/K/mol. Conveniently, we also have a way to test the volumes predicted by the model of \cite{DKS2013} at around 13 GPa. The experimentally-derived slope of the coesite melting curve is essentially zero at the triple point with stishovite at 13.7 GPa \citep{ZLGHF1993}. Therefore, from the Clausius-Clapeyron relationship $dT/dP = \Delta V / \Delta S$ the volumes of coesite and SiO$_2$ melt must be zero at this point. Unfortunately, the MD-derived SiO$_2$ melt model has its density crossover with coesite about 4 GPa lower. Therefore, we adjusted the bulk modulus and reference pressure to satisfy this constraint, and the volume at zero pressure. The adjustments required here may not be a failing of the MD simulations; it is likely that the coordination changes in the SiO$_2$ melt require a higher order Birch Murnaghan EoS than that used by \cite{DKS2013}. Our adjusted model is probably only accurate in a range of about $\pm$5 GPa around the triple point. The experimental data can also be used to judge the accuracy of the entropy of the SiO$_2$ melt model. The MD-derived entropy of melting is $\sim$40 J/K/mol at the triple point. This value is in perfect agreement with the value derived from the Clapeyron slope of the stishovite liquidus \citep{SL1995}. We note that the slope recently proposed by \cite{Millotetal2015} is much shallower, but their slope is a two parameter Simon's equation fit based on the triple point temperature at 13.7 GPa and the melting point at 500 GPa, and fits the data at 15--100 GPa \citep{SL1995, LAM1983} rather poorly. A Simon's equation fit is probably inappropriate over such extreme pressure ranges, especially where the melting point may be rather sensitive to changes in cation coordination.

\subsubsection{Solution model}
A subregular solution model was adopted to describe the free energy of intermediate-composition MgO-SiO$_2$ melts. The two interaction parameters W$_{MS}$ and W$_{SM}$ at 13 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions \citep{PG1990, PW1993}. We assumed an ideal entropic contribution to mixing. The pressure dependencies were fit to the melting curves. It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower. However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).

@@

\begin{equation}
p' = \frac{p_{H_2O}}{p_{H_2O}  + r \left(1.-p_{H_2O} \right)} 
\end{equation}
\begin{eqnarray}
X_O = 1 - p' - \frac{ 0.5 - \sqrt{ 0.25 - \frac{K_{(1)}-4}{K_{(1)}} \left(p'-p'^2 \right)}}{\frac{K_{(1)}-4}{K_{(1)}} }\\
X_{H_2O} = X_O + 2 p' - 1 \\
X_{OH} = 1 - X_O - X_{H_2O}
\end{eqnarray}
\noindent $p'$ is the molar proportion of H$_2$O where the silicate is normalised on a one oxygen basis (rather than a one-cation basis). We assume that all bridging oxygens are equally likely to be protonated. The number of oxygens per formula unit $r$ thus increases monotonously from 1 to 2 across the MgO-SiO$_2$ binary ($r_{\textrm{per}}$ = 1, $r_{\textrm{fo, wad, rw}}$ = 4/3, $r_{\textrm{oen, cen}}$ = 3/2, and $r_{\textrm{coe, stv}}$ = 2). The ideal activities of the anhydrous and water components can then be calculated:
\begin{eqnarray}
a^{\textit{s}}_{\textit{ideal melt}} =X_O^r \\
a^{H_2O}_{\textit{ideal melt}} = X_{H_2O}
\end{eqnarray}

@@

\clearpage
\section{Results}
\subsection{Ex-situ observations}
Experimental details and results are shown in Table \ref{table:experiments}. Superliquidus runs revealed networks of dendritic crystals of brucite, forsterite, enstatite and stishovite. Below the liquidus, solid phases tended to aggregate at the cold end of the capsules. Periclase formed equant crystals often associated with quench overgrowths which appear brighter in BSE images. Forsterite formed tabular crystals about 100 $\mu$m long. Enstatite crystals were much smaller, forming ca. 5 $\mu$m equant grains. Stishovite also formed equant to blocky grains, but crystals only separated from the melts in chambers with a high water content. In addition to the difference in texture and shape between quench crystals and liquidus phases, subliquidus phases typically contained many small inclusions of Pt-Rh oxides. A typical set of run products is shown in Figure \ref{fig:sem}.

\begin{figure}[h!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/assembly}
  \caption{The 18/11 octahedral assembly design used in this study.}
  \label{fig:assembly}
\end{figure}

\begin{landscape}
\begin{table}[ht!]
\caption{Experimental run conditions and run products determined by SEM/EDS. Compositions are listed in order of increasing molar H$_2$O content. Unused compositions for each run are marked by a dash. Minor solids are listed in brackets. Mineral abbreviations are as follows: br - brucite, per - periclase, co - hydroxychondrodite, en - clinoenstatite, s - stishovite. Melts are present in all run products.}
\label{table:experiments}
\begin{tabular}{llllccccc}
Expt \# & P (GPa) & T ($^{\circ}$C) & t (min) & brucite+PtAc & forsterite+PtAcid & enstatite+PtAcid & quartz+PtAcid & br+q \\
Z1063 & 13 & 1200 & 40 & -/br/br/br/L & co/fo/fo/(fo) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1085 & 13 & 1300 & 40 & -/-/per/L/L & -/fo/(en)/L & -/-/-/- & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1079 & 13 & 1300 & 35 & -/per/per/L/L & fo/fo/en/(en) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1058 & 13 & 1400 & 30 & -/per/(per)/L/L & fo/(fo)/L/L & en/en/en/(en) & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1060 & 13 & 1500 & 20 & -/-/-/-/- & (fo)/L/L/L & en/en/L/L & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1140 & 13 & 1600 & 10 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/s/(s) & s+en/en/en/en \\
Z1084 & 13 & 1650 & 10 & -/-/-/-/- & -/-/-/- & L/L/-/- & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1207 & 13 & 1700 & 8 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/(s)/L & s+en/(en)/L/L \\
Z1209 & 13 & 1800 & 5 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/(s?)/L/L/L & s/L/L/L \\
Z1091 & 13 & 1900 & 5 & (per)/-/-/-/- & -/-/-/- & -/-/-/- & -/-/-/L/-/L/L/L & L/L/-/-
\end{tabular}
\end{table}
\end{landscape}

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/Z1058_1400C_13GPa_fo}
  \caption{BSE image of typical run products. Experimental chambers from Experiment Z1058, conducted at 1400 $^{\circ}$C and 13 GPa. Chambers have compositions along the forsterite-water binary, with water contents increasing from left to right.}
  \label{fig:sem}
\end{figure}
